---
title: "4 - network analysis"
author: "Clint McKenna"
date: "9/15/2020"
output: html_document
---


## data prep

```{r}
# load packages
library(tidyverse)

# read in data from csv
tennis <- read.csv('../data/raw/wta_matches_2015.csv')
collegefb <- read.csv('../data/raw/CFBeattendance.csv')
netflix <- read.csv('../data/raw/netflix_titles.csv')
books <- read.csv('../data/raw/books.csv')


# custom functions
select <- dplyr::select
# ggplot2 alpha (sometimes gets a conflict depending on other packages)
alpha <- scales::alpha

# color palette
UMcolors <- c('#ffcb05', '#00274c', '#407EC9', '#9A3324', '#702082', '#D86018', '#00B2A9')

```


## creating edgelist

This will be somewhat related to the network stuff we did during the school year. For starters, we will look at the college football data. Normally we would want to have a complete dataset of all games, but this will be practice.

As a reminder, one way of organizing relational data is an edgelist. Typically if we have a directed edgelist
Its pretty close already since there is already a Team and Opponent column. We will organize it to have a "sender" col

```{r}
edgelist <- 


```

You can see there is a clear ceiling where Attendance can't exceed capacity. Here is the base of plotting correlation with ggplot. geom_point() adds "points" for datum.

```{r}
# plot
ggplot(collegefb, 
       aes(x = Stadium.Capacity,
           y = Attendance)) +
  geom_point()

```

Reminder that ?function_name() can check the help documentation in Rstudio

```{r}
# check arguments
?geom_point()

```

Like some of the previous functions, we can do things like adjust the size, color, alpha, etc.
```{r}
# plot
ggplot(collegefb, 
       aes(x = Stadium.Capacity,
           y = Attendance)) +
  geom_point(size = 2, color = 'red', alpha=  .05)

```

This is all pretty straightforward, but one of the benefits of using ggplot is adding in multiple layers. One thing I do a lot is integrate a fitted model into the graph. So if you have two continuous variables, this would estimate a regression equation.

I do this with stat_smooth(). Also takes several arguments to style the line.

```{r}

# plot
ggplot(collegefb, 
       aes(x = Stadium.Capacity,
           y = Attendance)) +
  geom_point(size = 2, color = UMcolors[1], alpha=  .15) +
  stat_smooth(method = 'lm',       # linear regression (can also do glm and other models)
              level = .95,         # 95% confidence intervals
              color = UMcolors[2], # color of line
              size = 1,            # size of line
              alpha = .75)         # transparency

```

This is also supposed to show the localized confidence intervals around the fitted line. I was confused why it wasn't showing up before I realized that it is just really small because there are so many observations: https://stackoverflow.com/a/49697067

We can recreate this with extremely stringent confidence intervals.

```{r}
# plot
ggplot(collegefb, 
       aes(x = Stadium.Capacity,
           y = Attendance)) +
  geom_point(size = 2, color = UMcolors[1], alpha=  .15) +
  stat_smooth(method = 'lm',       
              level = .999, 
              color = UMcolors[2],
              size = .25,           
              alpha = .75)      

```
Now you can see it (very thin) towards the outliers.


## group data

Now, let's look at integrating group (categorical predictors). This would be similar to testing an interaction and plotting it with regression. For instance:

```{r}
# model
# the asterisk means "add these variables and all interactions among them" to the model
m <- lm(Attendance ~ Stadium.Capacity * Tailgating, data = collegefb)
summary(m)


```

Because the interaction term is significant, we should see a different relationship strength for tailgating games vs. non-tailgating games.

Here we add group argument and move color to aes(), and remove color from geom_point(). For group and color, we specify the variable that is to be colored.

```{r}
# plot
ggplot(collegefb, 
       aes(x = Stadium.Capacity,
           y = Attendance,
           group = Tailgating,
           color = Tailgating)) +
  geom_point(size = 2, alpha=  .15) +
  stat_smooth(method = 'lm',       
              level = .95, 
              color = UMcolors[2],
              size = .25,           
              alpha = .75)      

```

Now to style the colors, we can use scale_color_manual() and scale_fill_manual()

```{r}
# plot
ggplot(collegefb, 
       aes(x = Stadium.Capacity,
           y = Attendance,
           group = Tailgating,
           color = Tailgating)) +
  geom_point(size = 2, alpha=  .15) +
  stat_smooth(method = 'lm',       
              level = .95, 
              size = .25,           
              alpha = .75) +
  scale_color_manual(values = UMcolors[2:1]) +
  scale_fill_manual(values = UMcolors[2:1])

```

The slopes actually look very similar for the two groups. But Tailgated games seem to only occur for very large stadiums (or perhaps the Tailgating variable is only recorded for large stadiums). We can look at only big games by filtering only games in stadiums with a capactity above, lets say, 50,000.

I also added here an aes() argument to stat_smooth for styling the confidence interval ribbons around the fitted lines (last graph it came out grey because I didn't specify this)

```{r}
# filter data
largefb <- collegefb %>%
  filter(Stadium.Capacity > 50000)


# plot
ggplot(largefb, 
       aes(x = Stadium.Capacity,
           y = Attendance,
           group = Tailgating,
           color = Tailgating)) +
  geom_point(size = 2, alpha = .15) +
  stat_smooth(method = 'lm',       
              level = .95, 
              size = .25,           
              alpha = .25,
              aes(fill = Tailgating)) +
  scale_color_manual(values = UMcolors[2:1]) +
  scale_fill_manual(values = UMcolors[2:1])


```

Here, we can see that the slopes slightly increase at a greater rate for tailgated games. That is, attendance goes up for stadiums that allow tailgating, but that might be driven by those outliers at the far end. A quick look at the data suggests that Beaver Stadium at Penn State is both large and allows Tailgating.

## facets

Another quick concept I'll introduce here are facets. I don't think we did these yet. Basically if you have a categorical variable that you want to break up into multiple plots, you can add facet_wrap or facet_grid at the end.

Here we can just break out the Tailgating variable like so:

```{r}

# plot
ggplot(largefb, 
       aes(x = Stadium.Capacity,
           y = Attendance,
           group = Tailgating,
           color = Tailgating)) +
  geom_point(size = 2, alpha = .15) +
  stat_smooth(method = 'lm',       
              level = .95, 
              size = .25,           
              alpha = .25,
              aes(fill = Tailgating)) +
  scale_color_manual(values = UMcolors[2:1]) +
  scale_fill_manual(values = UMcolors[2:1]) +
  facet_grid(~ Tailgating)

```

This is especially handy for when I am looking at 3-way interactions. Imagine adding a 3rd variable to group the data by, like Conference:


```{r}

# plot
ggplot(largefb, 
       aes(x = Stadium.Capacity,
           y = Attendance,
           group = Tailgating,
           color = Tailgating)) +
  geom_point(size = 2, alpha = .15) +
  stat_smooth(method = 'lm',       
              level = .95, 
              size = .25,           
              alpha = .25,
              aes(fill = Tailgating)) +
  scale_color_manual(values = UMcolors[2:1]) +
  scale_fill_manual(values = UMcolors[2:1]) +
  facet_wrap(~ Conference)

```
As you can see, some of the smaller conferences don't even record Tailgating!




## PRACTICE

Short practice for this one. I want you to read in the netflix dataset, and filter it so that we are only looking at the following:
- movies 
- released 1980 or later
- ratings of only G, PG, PG-13, and R

I also want you to make a column for the duration of the movie that is numeric and just the number of minutes for the length of the movie.



```{r}
netflix1<- read_csv('../data/raw/netflix_titles.csv')
  
netflix1<-netflix1 %>%
  
  #filter movies, year, rating
  filter(type == 'Movie')%>%
  filter(release_year >= 1980) %>%
  filter(rating == 'G' | rating == 'PG' | rating == 'PG-13' | rating == 'R') %>%
  
  #make new column with minutes, didn't want to get rid of duration column completely...
  mutate(minutes = duration) %>%
  separate(minutes, into = c('minutes'), sep = ' ')

#change to integer
netflix1$minutes <- strtoi(netflix1$minutes)

```


Plot the relation of release year (x-axis) and movie duration (y-axis) using geom_point() and add a regression fit line.

```{r}

ggplot(netflix1, aes(x = release_year, y = minutes)) + 
  geom_point(color = UMcolors[1]) + 
  stat_smooth(method = 'lm', 
              level = .95, 
              color = UMcolors[2], 
              size = 1) + 
  labs(title = 'Movie Runtime vs. Release Year on Netflix', 
       x = 'Release Year', 
       y = 'Runtime (minutes)')




```

Plot again, but add in rating as a categorical factor, either as a facet or color. In either case, add a color scheme to the plot of your choice.

```{r}

#facet
ggplot(netflix1, aes(x = release_year, y = minutes)) + 
  geom_point(color = UMcolors[2]) +
  stat_smooth(method = 'lm', 
              level = .95, 
              color = UMcolors[1]) + 
  facet_wrap(~rating)


```






















